---
title: "Figure 4"
date: "`r Sys.Date()`"
author: "Alex Qin"
output: 
  html_document: 
    self_contained: true
    code_folding: hide
    code_download: true 
    toc: true
    toc_float: true
    number_sections: true
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(ggsurvfit)
library(ggrepel)
library(spicyR)
library(Statial)
library(dplyr)
library(ggplot2)
library(SingleCellExperiment)
```


# Preprocessing - Sqrt Transform

```{r normalize-matrix-sqrt}
spe_Schurch_2020 <- readRDS("spe_Schurch_2020.rds")

sqrtSCE <- spe_Schurch_2020

assay(sqrtSCE, "intensities") <- apply(assay(spe_Schurch_2020, "intensities"), 2, sqrt)
# sqrtSCE <- trim99(sqrtSCE, markers = rownames(sqrtSCE), imageID = "imageID")

#Check if it worked
head(t(assay(sqrtSCE)))[,1:6]
head(t(assay(spe_Schurch_2020)))[,1:6]

schurchSCE <- sqrtSCE

# tmp1 <- schurchSCE |>
#   colData() |>
#   as.data.frame() |>
#   group_by(imageID, cellType) |>
#   summarise(cellCounts = n()) |>
#   filter(cellCounts >= 20) |>
#   select(imageID, cellType)
# 
# tmp2 <- schurchSCE |>
#   colData() |>
#   as.data.frame() |>
#   tibble::rownames_to_column("rownames") |>
#   inner_join(tmp1)
# 
# tmp <- schurchSCE[,tmp2$rownames]
# schurchSCE <- tmp

```

# Filtering down cell types

```{r}
schurchSCE$cellType |> table() |> sort()

schurchSCE <- schurchSCE[,schurchSCE$cellType != "dirt"]

colData(schurchSCE) = schurchSCE |> 
  colData() |>
  as.data.frame() |>
  mutate(cellType = case_when(
    cellType == "cd163_macrophages" ~ "macrophages",
    cellType == "cd4_t_cells_gata3" ~ "cd4_t_cells",
    cellType == "cd68_macrophages_gzm_b" ~ "macrophages",
    cellType == "cd3_t_cells" ~ "remove",
    cellType == "cd11c_d_cs" ~ "dendritic",
    cellType == "cd11b_cd68_macrophages" ~ "macrophages",
    cellType == "tumor_cells_immune_cells" ~ "remove",
    cellType == "cd68_macrophages" ~ "macrophages",
    cellType == "immune_cells_vasculature" ~ "remove",
    cellType == "undefined" ~ "remove",
    cellType == "cd68_cd163_macrophages" ~ "macrophages",
    cellType == "cd4_t_cells_cd45ro" ~ "cd4_t_cells",
    TRUE ~ cellType
  )) |>
  DataFrame()

schurchSCE <- schurchSCE[,schurchSCE$cellType != "remove"]
```

# Preprocessing - Distance

```{r, eval = TRUE}
set.seed(51773)
schurchSCE <- getDistances(schurchSCE,
                    maxDist = 200,
                    nCores = 20)

stateChangesSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "distances",
  minCells = 20,
  nCores = 40)

schurchSCE <- calcContamination(schurchSCE)

stateChangesCorrectedSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "distances",
  minCells = 20,
  nCores = 40,
  contamination = TRUE)

```

# Preprocessing - Abundance

```{r, eval = TRUE}
set.seed(51773)
schurchSCE <- getAbundances(schurchSCE, 
                          r = 200,
                          nCores = 20)

stateChangesAbuSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "abundances",
  minCells = 20,
  nCores = 40)

schurchSCE <- calcContamination(schurchSCE)

stateChangesCorrectedAbuSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "abundances",
  minCells = 20,
  nCores = 40,
  contamination = TRUE)
```

# Survival Preprocessing

```{r}
survData = schurchSCE |>
    na.omit() |>
    colData() |> 
    as.data.frame() |> 
    select(imageID, patient, os, os_censor, dfs, dfs_censor) |>
    unique()

# survData$new_oscensor <- ifelse(survData$os < 48, 1, survData$os_censor)

schurchSurv = Surv(survData$os, survData$os_censor)
names(schurchSurv) = survData$imageID
# stateByPatient <- stateChangesSchurch %>% na.omit()
# nrow(stateByPatient)
# stateByPatient <- stateByPatient %>% filter(imageID %in% survDataByPatient$imageID)
# nrow(stateByPatient)
# Creating survival vector


schurchSurvDF = Surv(survData$dfs, survData$dfs_censor)
names(schurchSurvDF) = survData$imageID


survData2 <- schurchSCE %>%
  colData() %>%
  as.data.frame() %>%
  select(imageID, patient, os, os_censor, dfs, dfs_censor) %>%
  unique %>%
  group_by(patient) %>%
  dplyr::slice(1) %>%
  ungroup()

# Filtering minCells for our calcStateChanges excludes image reg057_A (due to low number of cells), so we've taken reg057_B of the same patient for our survival object.

survData2 <- rbind(survData2, survData[survData$imageID == "reg057_B",])
survData2 <- survData2[survData2$imageID != "reg057_A",]

schurchSurv2 = Surv(survData2$os, survData2$os_censor)
names(schurchSurv2) = survData2$imageID


```

## For you to remember later: Figures 4A and 4B in the original SpatioMark manuscript is only reproducible when you set minCells = 0 for `calcStateChanges`. The default setting of minCells is minCells = 20. Otherwise the CD5 expression in CD8 T cells near Tumours does not come through for the survival plots, funnily enough.

If we do minCells = 100 - we have 33 significant relationships by FDR, but CD5 in CD8 T cells drops out, which is not a big deal I suppose. 

Will want to cross-check with Ellis what he thinks is the best way forward. I would just go for the minCells = 100 with the 33 relationships and change both the volcano plot and the survival plots. 

# Figure 4A (figuregen branch of Statial)

## Filtering by setting SD > 2

```{r}
stateMat <- prepMatrix(na.omit(stateChangesAbuSchurch), column = "tval")

# stateMat <- prepMatrix(stateChangesSchurch, column = "coef")

stateMat <- stateMat[names(schurchSurv2), ]
stateMat[is.na(stateMat)] <- 0
stateMat <- stateMat[,colMeans(abs(stateMat)>0.0001)>0.95]


survivalResults <- spicyR::colTest(stateMat, schurchSurv2, type = "survival")

head(survivalResults)
print(nrow(survivalResults[survivalResults$adjPval < 0.05 & survivalResults$coef > 0,]))
print(nrow(survivalResults[survivalResults$adjPval < 0.05 & survivalResults$coef < 0,]))
```

# Figure 4A

```{r}
survivalResults$mean <- survivalResults$coef
thresh <- 0.1
pvalThresh <- 0.05

survivalResults$diffexpressed <- "NO"
survivalResults$diffexpressed[survivalResults$mean > thresh & survivalResults$pval < pvalThresh] <- "UP"
survivalResults$diffexpressed[survivalResults$mean < -thresh & survivalResults$pval < pvalThresh] <- "DOWN"

survivalResults$combined <- rownames(survivalResults)
  
tmp1 <- survivalResults %>%
  filter(diffexpressed == "UP", adjPval < 0.05)
# tmp1$weightAvg <- tmp1$mean * tmp1$adjPval
# tmp1 <- head(tmp1[order(tmp1$weightAvg), "combined"], 7)
tmp1 <- head(tmp1[order(tmp1$adjPval), "combined"], 5)

tmp3 <- survivalResults %>%
  filter(diffexpressed == "DOWN", adjPval < 0.05)
tmp3 <- head(tmp3[order(tmp3$adjPval), "combined"], 5)
# tmp3$weightAvg <- abs(tmp3$mean) * tmp3$adjPval
# tmp3 <- head(tmp3[order(tmp3$weightAvg), "combined"], 7)

tmp <- c(tmp1, tmp3)

survivalResults$delabel <- ifelse(survivalResults$combined %in% tmp, survivalResults$combined, NA)

survivalResults <- survivalResults[order(survivalResults$adjPval),]


surVol <- ggplot(survivalResults, aes(x = mean, y = -log10(pval), col = diffexpressed, label = delabel)) +
  geom_vline(xintercept = c(-0.1, 0.1), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point() +
  xlim(-1, 1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00")) +
  geom_label_repel(max.overlaps = Inf, size = 3, min.segment.length = unit(0.1, "lines")) +
  theme(legend.position = "none") 
  labs(x = expression("Coefficient"), y = expression("-log"[10]*"(p-value)"))
  # scale_x_continuous(breaks = seq(-0.5, 0.5, 0.2)) +
  # coord_cartesian(xlim = c(-0.55, 0.55))

surVol
  
if (TRUE) {
    pdf.options(width=8.3, height=5)
    pdf(file="figs/Figure 4A.pdf")
    print(surVol)
    dev.off()
}
```

# Figure 4B

```{r}
# Selecting the most significant relationship
survRelationship = stateMat[["tumor_cells__macrophages__na_k_at_pase"]]
survRelationship = ifelse(survRelationship < median(survRelationship), "Higher expression in close cells", "Lower expression in close cells")
    
# Plotting Kaplan-Meier curve
p1 <- survfit2(schurchSurv2 ~ survRelationship) |>
    ggsurvfit() +
    add_pvalue(x = 500, y = 0.2) +
    ggtitle("tumor_cells__macrophages__na_k_at_pase") +
    add_censor_mark(shape = 4, size = 2)
    # add_confidence_interval()

if (TRUE) {
    pdf.options(width=9, height=4.5)
    pdf(file="figs/Figure 4Bi.pdf")
    print(p1)
    dev.off()
}

# Selecting the most significant relationship
survRelationship = stateMat[["tumor_cells__smooth_muscle__cd45ra"]]
survRelationship = ifelse(survRelationship < median(survRelationship), "Higher expression in close cells", "Lower expression in close cells")
    
# Plotting Kaplan-Meier curve
p2 <- survfit2(schurchSurv2 ~ survRelationship) |>
    ggsurvfit() +
    add_pvalue(x = 500, y = 0.2) +
    ggtitle("tumor_cells__smooth_muscle__cd45ra") +
    # add_confidence_interval() + 
    # xlim(0, 100)+
    add_censor_mark(shape = 4, size = 2)
    # xlim(0,60)

if (TRUE) {
    pdf.options(width=9, height=4.5)
    pdf(file="figs/Figure 4Bii.pdf")
    print(p2)
    dev.off()
}
```

# Figure 4C

```{r, eval = TRUE}
### Distances

# Preparing features for Statial
stateMat <- prepMatrix(na.omit(stateChangesSchurch), column = "tval")

# Ensuring rownames of stateMat match up with rownames of the survival vector
stateMat <- stateMat[names(schurchSurv), ]

stateMat[is.na(stateMat)] <- 0
# for(i in 1:ncol(stateMat)){
#   stateMat[is.na(stateMat[,i]) | stateMat[,i] == 0, i] <- mean(stateMat[,i], na.rm = TRUE)
# }
# Remove some very small values
stateMat <- stateMat[,colMeans(abs(stateMat)>0.0001)>.8]

stateMatCor <- prepMatrix(na.omit(stateChangesCorrectedSchurch), column = "tval")

# Ensuring rownames of stateMatCor match up with rownames of the survival vector
stateMatCor <- stateMatCor[names(schurchSurv), ]

stateMatCor[is.na(stateMatCor)] <- 0

stateMatCor <- stateMatCor[,colMeans(abs(stateMatCor)>0.0001)>.8]

stateMatAbu <- prepMatrix(na.omit(stateChangesAbuSchurch), column = "tval")

# Ensuring rownames of stateMatAbu match up with rownames of the survival vector
stateMatAbu <- stateMatAbu[names(schurchSurv), ]

stateMatAbu[is.na(stateMatAbu)] <- 0

stateMatAbu <- stateMatAbu[,colMeans(abs(stateMatAbu)>0.0001)>.8]

stateMatAbuCor <- prepMatrix(na.omit(stateChangesCorrectedAbuSchurch), column = "tval")

# Ensuring rownames of stateMatAbuCor match up with rownames of the survival vector
stateMatAbuCor <- stateMatAbuCor[names(schurchSurv), ]

stateMatAbuCor[is.na(stateMatAbuCor)] <- 0

stateMatAbuCor <- stateMatAbuCor[,colMeans(abs(stateMatAbuCor)>0.0001)>.8]

print(ncol(stateMat))
print(ncol(stateMatCor))
print(ncol(stateMatAbu))
print(ncol(stateMatAbuCor))
```

```{r, eval=FALSE}

schurchFiltSCE <- schurchSCE[,schurchSCE$imageID %in% survData$imageID]

schurchFiltSCE <- lisaClust::lisaClust(schurchFiltSCE, k = 5)

cellTypeRegionMeans <- getMarkerMeans(schurchFiltSCE,
                              imageID = "imageID",
                              cellType = "cellType",
                              region = "region")

cellTypeMarkerMeans <- getMarkerMeans(schurchFiltSCE,
                              imageID = "imageID",
                              cellType = "cellType")

cellTypeProp <- getProp(schurchFiltSCE, 
                       feature = "cellType",
                       imageID = "imageID")
regionProp <- getProp(schurchFiltSCE, 
                       feature = "region",
                       imageID = "imageID")

spicyData <- spicyR::getPairwise(schurchFiltSCE, 
                                 BPPARAM = BiocParallel::MulticoreParam(workers = 40)) %>%
  as.data.frame()
```

## Checking association with survival or other metrics

```{r, include = FALSE, eval=FALSE}
cellTypeMarkerMeans <- getMarkerMeans(schurchFiltSCE,
                              imageID = "imageID",
                              cellType = "cellType")

MarkerMeans <- getMarkerMeans(schurchFiltSCE,
                              imageID = "imageID")

cellTypeProp <- getProp(schurchFiltSCE, 
                       feature = "cellType",
                       imageID = "imageID")

stateMat <- stateMat[,colMeans(abs(stateMat)>0.0001)>.8]

cellTypeMarkerMeans <- cellTypeMarkerMeans[names(schurchSurv2), ]
cellTypeProp <- cellTypeProp[names(schurchSurv2), ]
MarkerMeans <- MarkerMeans[names(schurchSurv2), ]

survivalResults1 <- colTest(cellTypeMarkerMeans, schurchSurv2, type = "survival")
survivalResults2 <- colTest(cellTypeProp, schurchSurv2, type = "survival")
survivalResults3 <- colTest(MarkerMeans, schurchSurv2, type = "survival")

head(survivalResults)
```

```{r, warning=FALSE}
set.seed(51773)

# featurelist_comb <- c(featureList, region_two = featureListregs[1])
schurchCV_OS = ClassifyR::crossValidate(
  measurements = featureList,
  outcome = schurchSurv2,
  classifier = "randomForest",
  selectionMethod  = "CoxPH",
  multiViewMethod = "none",
  nFolds = 3,
  nFeatures = 10,
  nRepeats = 20,
  nCores = 40
)

cind <- ClassifyR::performancePlot(schurchCV_OS,
  characteristicsList = list(x = "auto", fillColour = "Assay Name")
  ) + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_manual(values = c("#ff8080", "red", "#8080ff", "blue", "grey", "grey", "grey", "grey", "grey"))

schurchCV_OS$ab_abundances.randomForest.CoxPH@chosenFeatures[[1]]$feature
schurchCV_OS$aa_distances.randomForest.CoxPH@chosenFeatures[[1]]$feature
cind
if (TRUE) {
    pdf.options(width=8.3, height=5)
    pdf(file="figs/Figure 4Ci test.pdf")
    print(cind)
    dev.off()
}

```

# KerenSCE

```{r, eval = TRUE}
kerenSPE = SpatialDatasets::spe_Keren_2018()

spatialDat = spatialCoords(kerenSPE) |> 
  data.frame()

kerenSPE$x = spatialDat$x
kerenSPE$y = spatialDat$y

kerenSPE
```


```{r, eval = TRUE}
set.seed(51773)
kerenSPE <- getDistances(kerenSPE,
                    maxDist = 200,
                    nCores = 20)

stateChangesKeren <- calcStateChanges(
  cells = kerenSPE,
  type = "distances",
  minCells = 20,
  nCores = 40)

kerenSPE <- calcContamination(kerenSPE)

stateChangesCorrectedKeren <- calcStateChanges(
  cells = kerenSPE,
  type = "distances",
  minCells = 20,
  nCores = 40,
  contamination = TRUE)
```


```{r, eval = TRUE}
set.seed(51773)
kerenSPE <- getAbundances(kerenSPE,
                          r = 200,
                          nCores = 20)

stateChangesAbuKeren <- calcStateChanges(
  cells = kerenSPE,
  type = "abundances",
  minCells = 20,
  nCores = 40)

kerenSPE <- calcContamination(kerenSPE)

stateChangesCorrectedAbuKeren <- calcStateChanges(
  cells = kerenSPE,
  type = "abundances",
  minCells = 20,
  nCores = 40,
  contamination = TRUE)
```

```{r, eval = TRUE}
# Extracting survival data
survData <- kerenSPE |>
  colData() |>
  data.frame() |>
  filter(tumour_type != "cold") |> 
  select(imageID, Survival_days_capped., Censored) |>
  mutate(event = 1 - Censored) |>
  unique() 

# Creating survival vector
kerenSurv <- Surv(survData$Survival_days_capped, survData$event)
names(kerenSurv) <- survData$imageID

# Remove NAs
kerenSurv = kerenSurv[!is.na(kerenSurv)]
```

```{r, eval = TRUE}
stateMat <- prepMatrix(na.omit(stateChangesKeren), column = "tval")
stateMat <- stateMat[names(kerenSurv), ]
stateMat[is.na(stateMat)] <- 0
stateMat <- stateMat[,colMeans(abs(stateMat)>0.0001)>.8]


stateMatCor <- prepMatrix(na.omit(stateChangesCorrectedKeren), column = "tval")
stateMatCor <- stateMatCor[names(kerenSurv), ]
stateMatCor[is.na(stateMatCor)] <- 0
stateMatCor <- stateMatCor[,colMeans(abs(stateMatCor)>0.0001)>.8]


stateMatAbu <- prepMatrix(na.omit(stateChangesAbuKeren), column = "tval")
stateMatAbu <- stateMatAbu[names(kerenSurv), ]
stateMatAbu[is.na(stateMatAbu)] <- 0
stateMatAbu <- stateMatAbu[,colMeans(abs(stateMatAbu)>0.0001)>.8]


stateMatAbuCor <- prepMatrix(na.omit(stateChangesCorrectedAbuKeren), column = "tval")
stateMatAbuCor <- stateMatAbuCor[names(kerenSurv), ]
stateMatAbuCor <- stateMatAbuCor[,colMeans(abs(stateMatAbuCor)>0.0001)>.8]

```

```{r, eval = TRUE}
set.seed(51773)
kerenFiltSCE <- kerenSPE[,kerenSPE$imageID %in% survData$imageID]

kerenFiltSCE <- lisaClust::lisaClust(kerenFiltSCE, k = 5)

cellTypeRegionMeans <- getMarkerMeans(kerenFiltSCE,
                              imageID = "imageID",
                              cellType = "cellType",
                              region = "region")

cellTypeMarkerMeans <- getMarkerMeans(kerenFiltSCE,
                              imageID = "imageID",
                              cellType = "cellType")

cellTypeProp <- getProp(kerenFiltSCE,
                       feature = "cellType",
                       imageID = "imageID")
regionProp <- getProp(kerenFiltSCE,
                       feature = "region",
                       imageID = "imageID")

spicyData <- spicyR::getPairwise(kerenFiltSCE,
                                 BPPARAM = BiocParallel::MulticoreParam(workers = 40)) %>%
  as.data.frame()
```

```{r, warning=FALSE, eval=TRUE}
set.seed(51773)
featureList <- list(
                    aa_distances = stateMat,
                    aa_distancesCorrected = stateMatCor,
                    ab_abundances = stateMatAbu,
                    ab_abundancesCorrected = stateMatAbuCor,
                    ac_regionMarkerMeans = cellTypeRegionMeans,
                    ad_cellTypeProp = cellTypeProp,
                    ae_regionProp = regionProp,
                    af_spicy = spicyData
                    )

# Ensure the rownames of the features match the order of the survival vector
featureList <- lapply(featureList, function(x)x[names(kerenSurv),])


kerenCV_OS = ClassifyR::crossValidate(
  measurements = featureList,
  outcome = kerenSurv,
  classifier = "randomForest",
  selectionMethod  = "CoxPH",
  multiViewMethod = "none",
  nFolds = 5,
  nFeatures = 10,
  nRepeats = 20,
  nCores = 40
)

# ClassifyR::performancePlot(kerenCV_OS,
#   characteristicsList = list(x = "auto", fillColour = "Assay Name")
#   ) +
#   theme(legend.position = "none") +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
#   # scale_fill_tableau()


cind <- ClassifyR::performancePlot(kerenCV_OS,
  characteristicsList = list(x = "auto", fillColour = "Assay Name")
  ) + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_manual(values = c("#ff8080", "red", "#8080ff", "blue", "grey", "grey", "grey", "grey"))

# schurchCV_OS$ab_abundances.randomForest.CoxPH@chosenFeatures[[1]]$feature
# schurchCV_OS$aa_distances.randomForest.CoxPH@chosenFeatures[[1]]$feature

if (TRUE) {
    pdf.options(width=8.3, height=5)
    pdf(file="figs/Figure 4C.pdf")
    print(cind)
    dev.off()
}

```

