---
title: "fig3"
output: html_document
date: "2024-01-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(ggsurvfit)
library(ggrepel)
library(spicyR)
library(Statial)
library(dplyr)
library(ggplot2)
library(SingleCellExperiment)
```

# Preprocessing - Sqrt Transform

```{r normalize-matrix-sqrt}

sqrtSCE <- spe_Schurch_2020

assay(sqrtSCE, "intensities") <- apply(assay(spe_Schurch_2020, "intensities"), 2, sqrt)
# sqrtSCE <- trim99(sqrtSCE, markers = rownames(sqrtSCE), imageID = "imageID")

#Check if it worked
head(t(assay(sqrtSCE)))[,1:6]
head(t(assay(schurchSCE)))[,1:6]

schurchSCE <- sqrtSCE
```

# Preprocessing - Distance

```{r}
schurchSCE <- getDistances(schurchSCE,
                    maxDist = 200,
                    nCores = 20)

stateChangesSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "distances",
  minCells = 100,
  nCores = 40)

schurchSCE <- calcContamination(schurchSCE)

stateChangesCorrectedSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "distances",
  nCores = 40,
  contamination = TRUE)

```
# Preprocessing - Abundance

```{r}
schurchSCE <- getAbundances(schurchSCE, 
                          r = 200,
                          nCores = 20)

stateChangesAbuSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "abundances",
  minCells = 100,
  nCores = 40)

schurchSCE <- calcContamination(schurchSCE)

stateChangesCorrectedAbuSchurch <- calcStateChanges(
  cells = schurchSCE,
  type = "abundances",
  nCores = 40,
  contamination = TRUE)
```
# Filtering down cell types

```{r, eval = FALSE}
schurchSCE$cellType |> table() |> sort()

schurchSCE <- schurchSCE[,schurchSCE$cellType != "dirt"]

colData(schurchSCE) = schurchSCE |> 
  colData() |>
  as.data.frame() |>
  mutate(cellType = case_when(
    cellType == "cd163_macrophages" ~ "macrophages",
    cellType == "cd4_t_cells_gata3" ~ "cd4_t_cells",
    cellType == "cd68_macrophages_gzm_b" ~ "macrophages",
    cellType == "cd3_t_cells" ~ "remove",
    cellType == "cd11c_d_cs" ~ "dendritic",
    cellType == "cd11b_cd68_macrophages" ~ "macrophages",
    cellType == "tumor_cells_immune_cells" ~ "remove",
    cellType == "cd68_macrophages" ~ "macrophages",
    cellType == "immune_cells_vasculature" ~ "remove",
    cellType == "undefined" ~ "remove",
    cellType == "cd68_cd163_macrophages" ~ "macrophages",
    cellType == "cd4_t_cells_cd45ro" ~ "cd4_t_cells",
    TRUE ~ cellType
  )) |>
  DataFrame()

schurchSCE <- schurchSCE[,schurchSCE$cellType != "remove"]
```

# Figure 3A-C (figuregen branch of Statial)

```{r}
plotStateChanges(schurchSCE,
                 image = "reg005_A",
                 from = "macrophages",
                 to = "vasculature",
                 marker = "cd34",
                 type = "distances",
                 purity_choice = "vasculature",
                 transform = "sqrt"
                 )$scatter

plotStateChanges(schurchSCE,
                 image = "reg005_A",
                 from = "macrophages",
                 to = "vasculature",
                 marker = "cd34",
                 type = "distances",
                 purity_choice = "vasculature",
                 transform = "sqrt"
                 )$contam_image

plotStateChanges(schurchSCE,
                 image = "reg007_B",
                 from = "macrophages",
                 to = "smooth_muscle",
                 marker = "a_sma",
                 type = "distances",
                 purity_choice = "smooth_muscle",
                 transform = )
## Dim = 8.54 * 5 Landscape

```

# Figure 3D

```{r}
ggplot(merged, aes(x = tval, y = tval_cor)) +
  ggrastr::geom_point_rast() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  theme_classic() +
  xlim(-30, 30) +
  ylim(-30, 30)
```


```{r}
merged <- inner_join(stateChangesSchurch, stateChangesCorrectedSchurch, by = c("imageID", "primaryCellType", "otherCellType", "marker"), suffix = c("", "_cor"))

merged$fdrDiff <- merged$fdr_cor/merged$fdr
merged$signfdr <- ifelse(merged$tval > 0, -log10(merged$fdr), -(-log10(merged$fdr)))
merged$signfdr_cor <- ifelse(merged$tval_cor > 0, -log10(merged$fdr_cor), -(-log10(merged$fdr_cor)))

merged$signpval <- ifelse(merged$tval > 0, -log10(merged$pval), -(-log10(merged$pval)))
merged$signpval_cor <- ifelse(merged$tval_cor > 0, -log10(merged$pval_cor), -(-log10(merged$pval_cor)))

merged$highlight <- "NO"
merged$highlight[merged$imageID == "reg065_A" & merged$marker == "cd45" & merged$primaryCellType == "tumor_cells" & merged$otherCellType == "cd68_cd163_macrophages"] <- "YES"

ggplot(data = merged, aes(x = signpval, y = signpval_cor)) +
  geom_point() +
  # geom_hex(bins = 50, alpha = 0.7) +
  # scale_fill_viridis_c(trans = "log", limit = c(1, NA)) +
  xlim(-50, 50) +
  ylim(-50, 50) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(log10(0.05), -log10(0.05)), linetype = "dashed", color = "red") +
  geom_hline(yintercept = c(log10(0.05), -log10(0.05)), linetype = "dashed", color = "red") +
  theme_classic()
```

# Figure 3E

```{r}
testCorrection <- function(cellTypeMarkers, # nolint
                           stateChanges,
                           stateChangesCorrected,
                           stateChanges2,
                           stateChanges2Corrected,
                           # stateChanges3,
                           # stateChanges3Corrected,
                           # stateChanges4,
                           # stateChanges4Corrected,
                           xlim = 100,
                           ylim = 10000,
                           title = "ROC",
                           unbalanced = FALSE
                           ) {
  cellTypeMarkers <- cellTypeMarkers

  values = c("#ff8080", "red", "#8080ff", "blue")
  names(values) <- c("dist", "dist_cor", "abu", "abu_cor")
  
  stateChanges <- stateChanges[!is.na(stateChanges$pval),]
  stateChangesCorrected <- stateChangesCorrected[!is.na(stateChangesCorrected$pval),]
  stateChanges2 <- stateChanges2[!is.na(stateChanges2$pval),]
  stateChanges2Corrected <- stateChanges2Corrected[!is.na(stateChanges2Corrected$pval),]
  
  if(unbalanced == TRUE) {
    
    tmp <- unite(stateChanges, "new", imageID:marker)
    stateChanges$united <- tmp$new

    tmp <- unite(stateChangesCorrected, "new", imageID:marker)
    stateChangesCorrected$united <- tmp$new
    
    tmp <- unite(stateChanges2, "new", imageID:marker)
    stateChanges2$united <- tmp$new
    
    tmp <- unite(stateChanges2Corrected, "new", imageID:marker)
    stateChanges2Corrected$united <- tmp$new
    
    stateChanges <- stateChanges[stateChanges$united %in% stateChanges2$united,]
    stateChangesCorrected <- stateChangesCorrected[stateChangesCorrected$united %in% stateChanges2Corrected$united,]
    
  }
  

  stateChanges <- stateChanges[order(stateChanges$pval),]
  stateChangesCorrected <- stateChangesCorrected[order(stateChangesCorrected$pval),]
  # 
  stateChanges2 <- stateChanges2[order(stateChanges2$pval),]
  stateChanges2Corrected <- stateChanges2Corrected[order(stateChanges2Corrected$pval),]
  # 
  # stateChanges3 <- stateChanges3[order(stateChanges3$fdr),]
  # stateChanges3Corrected <- stateChanges3Corrected[order(stateChanges3Corrected$fdr),]
  # 
  # stateChanges4 <- stateChanges4[order(stateChanges4$fdr),]
  # stateChanges4Corrected <- stateChanges4Corrected[order(stateChanges4Corrected$fdr),]

  # print(head(stateChanges))
  # print(head(stateChangesCorrected))
  # print(head(stateChanges2))
  # print(head(stateChanges2Corrected))

  # stateChanges <- stateChanges %>% filter(fdr < 0.05)
  # stateChangesCorrected <- stateChangesCorrected %>% filter(fdr < 0.05)
  # stateChanges2 <- stateChanges2 %>% filter(fdr < 0.05)
  # stateChanges2Corrected <- stateChanges2Corrected %>% filter(fdr < 0.05)
  # stateChanges3 <- stateChanges3 %>% filter(fdr < 0.05)
  # stateChanges3Corrected <- stateChanges3Corrected %>% filter(fdr < 0.05)
  # stateChanges4 <- stateChanges4 %>% filter(fdr < 0.05)
  # stateChanges4Corrected <- stateChanges4Corrected %>% filter(fdr < 0.05)



  df <- rbind(data.frame(TP =cumsum(!stateChanges$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges$fdr < 0.05 ~ "significant"), type = "dist"),
              data.frame(TP =cumsum(!stateChangesCorrected$marker %in% cellTypeMarkers),
                        FP = cumsum(stateChangesCorrected$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChangesCorrected$fdr < 0.05 ~ "significant"), type = "dist_cor"),
              data.frame(TP =cumsum(!stateChanges2$marker %in% cellTypeMarkers),
                        FP = cumsum(stateChanges2$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges2$fdr < 0.05 ~ "significant"), type = "abu"),
              data.frame(TP =cumsum(!stateChanges2Corrected$marker %in% cellTypeMarkers),
                        FP = cumsum(stateChanges2Corrected$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges2Corrected$fdr < 0.05 ~ "significant"), type = "abu_cor"))
              # data.frame(TP =cumsum(!stateChanges3$marker %in% cellTypeMarkers), 
              #           FP = cumsum(stateChanges3$marker %in% cellTypeMarkers),
              #           fdr = case_when(stateChanges3$fdr < 0.05 ~ "significant"), type = "contamF1All"),
              # data.frame(TP =cumsum(!stateChanges3Corrected$marker %in% cellTypeMarkers), 
              #           FP = cumsum(stateChanges3Corrected$marker %in% cellTypeMarkers), 
              #           fdr = case_when(stateChanges3Corrected$fdr < 0.05 ~ "significant"), type = "contamF1Def"),
              # data.frame(TP =cumsum(!stateChanges4$marker %in% cellTypeMarkers), 
              #           FP = cumsum(stateChanges4$marker %in% cellTypeMarkers), 
              #           fdr = case_when(stateChanges4$fdr < 0.05 ~ "significant"), type = "contamMainF1"),
              # data.frame(TP =cumsum(!stateChanges4Corrected$marker %in% cellTypeMarkers), 
              #           FP = cumsum(stateChanges4Corrected$marker %in% cellTypeMarkers),
              #           fdr = case_when(stateChanges4Corrected$fdr < 0.05 ~ "significant"), type = "noContam")) #nolint
  # print(head(df[df$type == "Corrected",]))
  # print(head(df[df$type == "None_BIDC",]))
  # print(tail(df[df$type == "None_BIDC",]))
  # print(head(df[df$type == "None_BIDC",]))
  
  
    df2 <- rbind(data.frame(TP =!stateChanges$marker %in% cellTypeMarkers, 
                        FP = stateChanges$marker %in% cellTypeMarkers,
                        fdr = case_when(stateChanges$fdr < 0.05 ~ "significant"), type = "dist"),
              data.frame(TP =!stateChangesCorrected$marker %in% cellTypeMarkers,
                        FP = stateChangesCorrected$marker %in% cellTypeMarkers,
                        fdr = case_when(stateChangesCorrected$fdr < 0.05 ~ "significant"), type = "dist_cor"),
              data.frame(TP =!stateChanges2$marker %in% cellTypeMarkers,
                        FP = stateChanges2$marker %in% cellTypeMarkers,
                        fdr = case_when(stateChanges2$fdr < 0.05 ~ "significant"), type = "abu"),
              data.frame(TP =!stateChanges2Corrected$marker %in% cellTypeMarkers,
                        FP = stateChanges2Corrected$marker %in% cellTypeMarkers,
                        fdr = case_when(stateChanges2Corrected$fdr < 0.05 ~ "significant"), type = "abu_cor"))

  slope <- nrow(stateChanges %>% filter(!marker %in% cellTypeMarkers))/nrow(stateChanges %>% filter(marker %in% cellTypeMarkers))
  slope <- slope %>% as.numeric
  p1 <- ggplot(df, aes(x = FP, y = TP, colour = type)) + 
        geom_line()+ labs(y = "Cell state marker", x = "Cell type marker") + 
        scale_colour_manual(values = values) +
        geom_abline(intercept = 0, slope = slope, linetype = "dashed", color = "#000000") +
        theme(legend.position = "none")


  p2 <- ggplot(df, aes(x = FP, y = TP, colour = type)) + 
        geom_line() +
        xlim(0,xlim) +
        ylim(0,ylim) +
        labs(y = "Cell state marker", x = "Cell type marker") + 
        scale_colour_manual(values = values) +
        geom_abline(intercept = 0, slope = slope, linetype = "dashed", color = "#000000") +
        theme(legend.position = "none")

  list(df, df2, p1, p2)
}
```

```{r}
testCorrection(cellTypeMarkers = c("cd3", "cd4", "cd8", "cd56", "cd11c", "cd68", "cd45", "cd20", "cd11b"),
               stateChanges = stateChangesSchurch,
               stateChangesCorrected = stateChangesCorrectedSchurch,
               stateChanges2 = stateChangesAbuSchurch,
               stateChanges2Corrected = stateChangesCorrectedAbuSchurch,
               # stateChanges3 = contamF1All,
               # stateChanges3Corrected = contamF1Def,
               # stateChanges4 = contamMainF1,
               # stateChanges4Corrected = stateChangesAbu,
               xlim = 100,
               ylim = 850,
               title = "",
               unbalanced = TRUE
              )[[4]]

auc_df <- testCorrection(cellTypeMarkers = c("cd3", "cd4", "cd8", "cd56", "cd11c", "cd68", "cd45", "cd20", "cd11b"),
               stateChanges = stateChangesSchurch,
               stateChangesCorrected = stateChangesCorrectedSchurch,
               stateChanges2 = stateChangesAbuSchurch,
               stateChanges2Corrected = stateChangesCorrectedAbuSchurch,
               # stateChanges3 = contamF1All,
               # stateChanges3Corrected = contamF1Def,
               # stateChanges4 = contamMainF1,
               # stateChanges4Corrected = stateChangesAbu,
               xlim = 100,
               ylim = 850,
               title = "",
               unbalanced = TRUE
              )[[1]]

auc_df2 <- testCorrection(cellTypeMarkers = c("cd3", "cd4", "cd8", "cd56", "cd11c", "cd68", "cd45", "cd20", "cd11b"),
               stateChanges = stateChangesSchurch,
               stateChangesCorrected = stateChangesCorrectedSchurch,
               stateChanges2 = stateChangesAbuSchurch,
               stateChanges2Corrected = stateChangesCorrectedAbuSchurch,
               # stateChanges3 = contamF1All,
               # stateChanges3Corrected = contamF1Def,
               # stateChanges4 = contamMainF1,
               # stateChanges4Corrected = stateChangesAbu,
               xlim = 100,
               ylim = 850,
               title = "",
               unbalanced = TRUE
              )[[2]]

auc_df |>
  group_by(type) |>
  summarise(total_positives = max(TP),
            total_negatives = max(FP))

auc_df$TPR <- auc_df$TP / 658067
auc_df$FPR <- auc_df$FP / 122076

plot(auc_df$FPR, auc_df$TPR, type = "l", col = "blue", xlab = "False Positive Rate", ylab = "True Positive Rate", main = "ROC Curve")

# Create a ROC object
roc_curve_dist <- roc(response = auc_df$TPR[auc_df$type == "dist"],
                      predictor = auc_df$FPR[auc_df$type == "dist"], 
                      plot = TRUE)

roc_curve_distcor <- roc(response = auc_df$TPR[auc_df$type == "dist_cor"], 
                         predictor = auc_df$FPR[auc_df$type == "dist_cor"],
                         plot = TRUE)

roc_curve_abu <- roc(response = auc_df$TPR[auc_df$type == "abu"], 
                     predictor = auc_df$FPR[auc_df$type == "abu"],
                     plot = TRUE)

roc_curve_abucor <- roc(response = auc_df$TPR[auc_df$type == "abu_cor"], 
                        predictor = auc_df$FPR[auc_df$type == "abu_cor"],
                        plot = TRUE)

# Calculate the partial AUC between FPR of 0 and 0.2 (adjust as necessary)
partial_auc <- auc(roc_curve_dist, partial.auc = c(0, 0.0008191618), partial.auc.focus = "specificity"
                   # , partial.auc.correct = TRUE
                   )
print(partial_auc)

partial_auc <- auc(roc_curve_distcor, partial.auc = c(0, 0.0008191618), partial.auc.focus = "specificity"
                   # , partial.auc.correct = TRUE
                   )
print(partial_auc)

partial_auc <- auc(roc_curve_abu, partial.auc = c(0, 0.0008191618), partial.auc.focus = "specificity"
                   # , partial.auc.correct = TRUE
                   )
print(partial_auc)

partial_auc <- auc(roc_curve_abucor, partial.auc = c(0, 0.0008191618), partial.auc.focus = "specificity"
                   # , partial.auc.correct = TRUE
                   )
print(partial_auc)

roc_curve <- roc(response = auc_df$TPR[auc_df$type == "dist_cor"], predictor = auc_df$FPR[auc_df$type == "dist_cor"])

# Calculate the partial AUC between FPR of 0 and 0.2 (adjust as necessary)
partial_auc <- auc(roc_curve, partial.auc = c(0, 0.001), partial.auc.focus = "sensitivity", partial.auc.correct = TRUE)
print(partial_auc)
# testCorrection(cellTypeMarkers = c("cd3", "cd4", "cd8", "cd56", "cd11c", "cd68", "cd45", "cd20", "cd11b"),
#                stateChanges = superListDistance[[1]],
#                stateChangesCorrected = superListDistance[[2]],
#                stateChanges2 = superListDistance[[7]],
#                stateChanges2Corrected = stateChangesSchurch,
#                # stateChanges3 = contamF1All,
#                # stateChanges3Corrected = contamF1Def,
#                # stateChanges4 = contamMainF1,
#                # stateChanges4Corrected = stateChangesAbu,
#                xlim = 100,
#                ylim = 850,
#                title = "",
#                unbalanced = TRUE
#               )[[2]]

plot.roc(aSAH$outcome, aSAH$s100b)
```
# ROCR

```{r}
library(ROCR)

auc_df <- auc_df |>
  group_by(type) |>
  tibble::rowid_to_column("ID")

# Since you have ranks, you can use them as scores
pred <- prediction(-auc_df$ID[auc_df$type == "dist"], auc_df2$TP[auc_df2$type == "dist"])

perf <- performance(pred, "tpr", "fpr")

plot(perf, col = "blue", lwd = 2, main = "ROC Curve from Rankings", xlim = c(0, 0.001), ylim = c(0, 0.001))
abline(a = 0, b = 1, lty = 2, col = "gray")

auc_perf <- performance(pred, measure = "auc", fpr.stop =0.001)
auc_value <- auc_perf@y.values[[1]]
legend("bottomright", legend = paste("AUC =", round(auc_value, 3)))

corrected_auc <- 0.5 * (1 + (auc_value - 0.5*0.001**2) / (0.001 - 0.5*0.001**2))
```


```{r}
# pAUC_thresh <- 0.003

for(pAUC_thresh in 1:2*0.001) {
  pred <- prediction(-auc_df$ID[auc_df$type == "dist"], auc_df2$TP[auc_df2$type == "dist"])

  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, col = "blue", lwd = 2, main = "ROC Curve from Rankings", xlim = c(0, pAUC_thresh), ylim = c(0, 2*pAUC_thresh))
  abline(a = 0, b = 1, lty = 2, col = "gray")
  
  auc_perf <- performance(pred, measure = "auc", fpr.stop =pAUC_thresh)
  auc_value <- auc_perf@y.values[[1]]
  legend("bottomright", legend = paste("AUC =", round(auc_value, 3)))
  
  corrected_auc <- 0.5 * (1 + (auc_value - 0.5*pAUC_thresh**2) / (pAUC_thresh - 0.5*pAUC_thresh**2))
  print(corrected_auc)
}

for(pAUC_thresh in 1:2*0.001) {
  pred <- prediction(-auc_df$ID[auc_df$type == "dist_cor"], auc_df2$TP[auc_df2$type == "dist_cor"])

  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, col = "blue", lwd = 2, main = "ROC Curve from Rankings", xlim = c(0, pAUC_thresh), ylim = c(0, 2*pAUC_thresh))
  abline(a = 0, b = 1, lty = 2, col = "gray")
  
  auc_perf <- performance(pred, measure = "auc", fpr.stop =pAUC_thresh)
  auc_value <- auc_perf@y.values[[1]]
  legend("bottomright", legend = paste("AUC =", round(auc_value, 3)))
  
  corrected_auc <- 0.5 * (1 + (auc_value - 0.5*pAUC_thresh**2) / (pAUC_thresh - 0.5*pAUC_thresh**2))
  print(corrected_auc)
}

```

The reason why the pAUC is so low is because the maximum pAUC is calculated by extending upwards. For example, if you set the `ylim = c(0,1)` and kept xlim as is, then your blue line pretty much blends in with the diagonal line, which is why having a low pAUC here close to 0.5 makes sense. 

```{r}
pred <- prediction(-auc_df$ID[auc_df$type == "dist"], auc_df2$TP[auc_df2$type == "dist"])
perf <- performance(pred, "tpr", "fpr")

# Extract TPR and FPR values
tpr_values <- perf@y.values[[1]]
fpr_values <- perf@x.values[[1]]

# Define the FPR range
fpr_min <- 0.0
fpr_max <- 0.001

# Select indices within the desired FPR range
indices <- which(fpr_values >= fpr_min & fpr_values <= fpr_max)

# Extract corresponding TPR and FPR
tpr_partial <- tpr_values[indices]
fpr_partial <- fpr_values[indices]


# Use the trapezoidal rule to approximate the area
partial_auc <- pracma::trapz(fpr_partial, tpr_partial)

# Compute the maximum possible area in the partial ROC space
max_partial_auc <- (fpr_max - fpr_min) * 1  # Since TPR can go up to 1

# Normalize the partial AUC
normalized_partial_auc <- partial_auc / max_partial_auc

print(paste("Partial AUC (FPR", fpr_min, "to", fpr_max, ") =", round(normalized_partial_auc, 5)))

```


```{r}
# Example: If lower ranks mean higher confidence

auc_df <- auc_df |>
  group_by(type) |>
  mutate(adjusted_score = max(ID) - ID + 1) |>
  mutate(scaled_score = adjusted_score/max(adjusted_score))

# roc_obj <- roc(response = c(rep(TRUE, nrow(auc_df2[auc_df2$type == "dist" & auc_df2$TP == TRUE,])), 
#                             rep(FALSE, nrow(auc_df2[auc_df2$type == "dist" & auc_df2$TP == FALSE,]))), 
#                predictor = auc_df$scaled_score[auc_df$type == "dist"]) |> plot()

roc1 <- roc(response = auc_df2$TP[auc_df2$type == "dist"], 
               predictor = auc_df$scaled_score[auc_df$type == "dist"],
               auc = TRUE,
               partial.auc = c(1, 0.9995),
               partial.auc.focus = "specificity"
               # partial.auc.correct = TRUE
               ) 
roc1
roc1 |>
    plot(xlim = c(1, 0.999), ylim = c(0, 0.001))

roc2 <- roc(response = auc_df2$TP[auc_df2$type == "dist_cor"], 
               predictor = auc_df$scaled_score[auc_df$type == "dist_cor"],
               auc = TRUE,
               partial.auc = c(1, 0.9995),
               partial.auc.focus = "specificity"
               # partial.auc.correct = TRUE
               # partial.auc.correct = TRUE
               ) 
roc2
roc2 |>
    plot(xlim = c(1, 0.999), ylim = c(0, 0.001))

roc.test(roc1, roc2, partial.auc=c(1, 0.999))

# roc_obj <- roc(response = auc_df2$TP[auc_df2$type == "dist_cor"], predictor = auc_df$adjusted_score[auc_df$type == "dist_cor"]) |>
#   plot(xlim = c(1, 0.999), ylim = c(0, 0.001))
# 
# plot(roc_obj, xlim = c(0, 1), ylim = c(0, 1), col = "blue", lwd = 2,
#      main = "Partial ROC Curve (FPR 0 to 0.2)")
# abline(a = 0, b = 1, lty = 2, col = "gray")
# 
# roc(aSAH$outcome, aSAH$age) |> plot()

roc1$auc
roc2$auc
```

```{r}
# Function to calculate pAUC relative to random using trapezoidal rule
pauc_relative_to_random <- function(fpr, tpr, fpr_min, fpr_max) {
  # Filter points within the specified FPR range
  within_range <- which(fpr >= fpr_min & fpr <= fpr_max)
  # Get the FPR and TPR values within the range
  fpr_subset <- fpr[within_range]
  tpr_subset <- tpr[within_range]
  # If necessary, add boundary points
  if (min(fpr_subset) > fpr_min) {
    fpr_subset <- c(fpr_min, fpr_subset)
    tpr_subset <- c(approx(fpr, tpr, xout = fpr_min)$y, tpr_subset)
  }
  if (max(fpr_subset) < fpr_max) {
    fpr_subset <- c(fpr_subset, fpr_max)
    tpr_subset <- c(tpr_subset, approx(fpr, tpr, xout = fpr_max)$y)
  }
  # Calculate the raw partial AUC using the trapezoidal rule
  raw_pauc <- 0
  for (i in 1:(length(fpr_subset) - 1)) {
    # Trapezoid area: 0.5 * (x2 - x1) * (y1 + y2)
    raw_pauc <- raw_pauc +
      0.5 * (fpr_subset[i+1] - fpr_subset[i]) * (tpr_subset[i] + tpr_subset[i+1])
  }
  # Calculate the maximum possible pAUC in the range [fpr_min, fpr_max]
  max_pauc <- fpr_max - fpr_min
  # Calculate the random pAUC (area under the diagonal line y = x)
  random_pauc <- 0.5 * (fpr_max - fpr_min)
  # Calculate pAUC relative to random
  if (max_pauc == random_pauc) {
    relative_pauc <- 0  # To avoid division by zero, set relative pAUC to 0
  } else {
    relative_pauc <- (raw_pauc - random_pauc) / (max_pauc - random_pauc)
  }
  return(relative_pauc)
}
# Example usage:
# Define some example FPR and TPR values (ROC curve points)
# Specify the partial FPR range
fpr_min <- 0
fpr_max <- 0.001
# Calculate the pAUC relative to random
relative_pauc_value <- pauc_relative_to_random(auc_df$FPR[auc_df$type == "dist"], auc_df$TPR[auc_df$type == "dist"], fpr_min, fpr_max)
cat("Partial AUC relative to random (between 0 and 1):", relative_pauc_value, "\n")
```





```{r}
pROC = function (truth, stat = NULL,p)
{
 if (!length(truth))
 return(NULL)
 truth <- as.numeric(as.logical(truth))
 if (!is.null(stat)) {
 if (length(stat) != length(truth))
 stop("lengths differ")
 truth[is.na(stat)] <- NA
 truth <- truth[order(stat, decreasing = TRUE)]
 }
 isna <- is.na(truth)
 if (any(isna))
 truth <- truth[!isna]
 use = cumsum(truth==0)<p
 sens <- cumsum(truth)/sum(truth)
 truth1 = truth[use]
 mean(sens[which(truth1 == 0)])
}

pROC(auc_df2$FP[auc_df2$type == "dist"], p = 100)

TPR = function(x)pROC((names(sort(x[names(x)%in%c(TRUEDE,NOTDE)]))%in%TRUEDE)*1,p = 10)

```


# Figure 3F
## Preprocessing
```{r}
library(CellSPA)

### BIDCell

XeBreast <- readRDS("/dskh/nobackup/biostat/projects/singlecell/CellSPA_analysis/results/xenium_breast/spe_objects/spe_BIDCell_withMetrics.rds")

colData(XeBreast) <- cbind(colData(XeBreast), spatialCoords(XeBreast))

tmp <- colData(XeBreast) %>% as.data.frame
tmp$scClassify <- gsub("[[:space:]+]", ".", tmp$scClassify)
colData(XeBreast) <- tmp %>% DataFrame()

assay(XeBreast, "logcounts") <- as.matrix(assay(XeBreast, "logcounts"))

XeBreast <- getDistances(XeBreast,
                        maxDist = 200,
                        imageID = "sample_id",
                        cellType = "scClassify",
                        nCores = 20,
                        spatialCoords = c("cell_centroid_x", "cell_centroid_y"))

XeBreast <- calcContamination(XeBreast, assay = "logcounts", cellType = "scClassify")

reducedDim(XeBreast, "contaminations")[["scClassify"]] <- NULL

stateChangesBID <- calcStateChanges(
  cells = XeBreast,
  type = "distances",
  assay = "logcounts",
  imageID = "sample_id",
  cellType = "scClassify",
  minCells = 100,
  nCores = 1
)

stateChangesBIDConta <- calcStateChanges(
  cells = XeBreast,
  type = "distances",
  assay = "logcounts",
  imageID = "sample_id",
  cellType = "scClassify",
  minCells = 100,
  nCores = 1,
  contamination = TRUE
)

### nuclei

XeBreast10x <- readRDS("/dskh/nobackup/biostat/projects/singlecell/CellSPA_analysis/results/xenium_breast/spe_objects/spe_10x_nuclei_withMetrics.rds")

colData(XeBreast10x) <- cbind(colData(XeBreast10x), spatialCoords(XeBreast10x))

tmp <- colData(XeBreast10x) %>% as.data.frame
tmp$scClassify <- gsub("[[:space:]+]", ".", tmp$scClassify)
colData(XeBreast10x) <- tmp %>% DataFrame()

assay(XeBreast10x, "logcounts") <- as.matrix(assay(XeBreast10x, "logcounts"))


XeBreast10x <- getDistances(XeBreast10x,
                        maxDist = 200,
                        imageID = "sample_id",
                        cellType = "scClassify",
                        nCores = 20,
                        spatialCoords = c("cell_centroid_x", "cell_centroid_y"))

stateChanges10x <- calcStateChanges(
  cells = XeBreast10x,
  type = "distances",
  # from = "B.Cells",
  # to = c("CD4.T", "CD8.T"),
  # marker = c("CD3D", "CD3E", "CD3G", "CD4", "CD8A", "CD8B"),
  imageID = "sample_id",
  cellType = "scClassify",
  assay = "logcounts",
  nCores = 1,
  minCells = 100)

XeBreast10x <- calcContamination(XeBreast10x, assay = "logcounts", cellType = "scClassify")

reducedDim(XeBreast10x, "contaminations")[["scClassify"]] <- NULL

stateChanges10xConta <- calcStateChanges(
  cells = XeBreast10x,
  type = "distances",
  # from = "B.Cells",
  # to = c("CD4.T", "CD8.T"),
  # marker = c("CD3D", "CD3E", "CD3G", "CD4", "CD8A", "CD8B"),
  imageID = "sample_id",
  cellType = "scClassify",
  assay = "logcounts",
  nCores = 1,
  minCells = 100,
  contamination = TRUE)

### Voronoi


VoronoiXe <- readRDS("/dskh/nobackup/biostat/projects/singlecell/CellSPA_analysis/results/xenium_breast/spe_objects/spe_Voronoi_withMetrics.rds")

colData(VoronoiXe) <- cbind(colData(VoronoiXe), spatialCoords(VoronoiXe))

tmp <- colData(VoronoiXe) %>% as.data.frame
tmp$scClassify <- gsub("[[:space:]+]", ".", tmp$scClassify)
colData(VoronoiXe) <- tmp %>% DataFrame()

assay(VoronoiXe, "logcounts") <- as.matrix(assay(VoronoiXe, "logcounts"))

VoronoiXe <- getDistances(VoronoiXe,
                        maxDist = 200,
                        imageID = "sample_id",
                        cellType = "scClassify",
                        nCores = 20,
                        spatialCoords = c("cell_centroid_x", "cell_centroid_y"))

VoronoiXe <- calcContamination(VoronoiXe, assay = "logcounts", cellType = "scClassify")

reducedDim(VoronoiXe, "contaminations")[["scClassify"]] <- NULL

stateChangesVOR <- calcStateChanges(
  cells = VoronoiXe,
  type = "distances",
  assay = "logcounts",
  imageID = "sample_id",
  cellType = "scClassify",
  minCells = 100,
  nCores = 1
)

stateChangesVORConta <- calcStateChanges(
  cells = VoronoiXe,
  type = "distances",
  assay = "logcounts",
  imageID = "sample_id",
  cellType = "scClassify",
  minCells = 100,
  nCores = 1,
  contamination = TRUE
)
```

## Defining functions
```{r}
testCorrection <- function(cellTypeMarkers, # nolint
                           stateChanges, # nolint
                           stateChangesCorrected,
                           stateChanges2,
                           stateChanges2Corrected,
                           stateChanges3,
                           stateChanges3Corrected,
                           xlim = 100,
                           ylim = 10000,
                           title = "what",
                           fdr = TRUE
                           ) {
  cellTypeMarkers <- cellTypeMarkers

  values = c("#ff8080", "#ff0000", "#8080ff", "#0000ff", "#00ff55", "#009933")
  names(values) <- c("None_BIDC", "Corrected_BIDC", "None_VOR", "Corrected_VOR", "None_10x", "Corrected_10x")

  stateChanges <- stateChanges[order(stateChanges$fdr),]
  stateChangesCorrected <- stateChangesCorrected[order(stateChangesCorrected$fdr),]

  stateChanges2 <- stateChanges2[order(stateChanges2$fdr),]
  stateChanges2Corrected <- stateChanges2Corrected[order(stateChanges2Corrected$fdr),]
  
  stateChanges3 <- stateChanges3[order(stateChanges3$fdr),]
  stateChanges3Corrected <- stateChanges3Corrected[order(stateChanges3Corrected$fdr),]

  # print(head(stateChanges))
  # print(head(stateChangesCorrected))
  # print(head(stateChanges2))
  # print(head(stateChanges2Corrected))
  if (fdr == TRUE) {
    stateChanges <- stateChanges %>% filter(fdr < 0.05)
    stateChangesCorrected <- stateChangesCorrected %>% filter(fdr < 0.05)
    stateChanges2 <- stateChanges2 %>% filter(fdr < 0.05)
    stateChanges2Corrected <- stateChanges2Corrected %>% filter(fdr < 0.05)  
    stateChanges3 <- stateChanges3 %>% filter(fdr < 0.05)
    stateChanges3Corrected <- stateChanges3Corrected %>% filter(fdr < 0.05)  
  }
  



  df <- rbind(data.frame(TP =cumsum(!stateChanges$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges$fdr < 0.05 ~ "significant"), type = "None_BIDC"),
              data.frame(TP =cumsum(!stateChangesCorrected$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChangesCorrected$marker %in% cellTypeMarkers), 
                        fdr = case_when(stateChangesCorrected$fdr < 0.05 ~ "significant"), type = "Corrected_BIDC"),
              data.frame(TP =cumsum(!stateChanges2$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges2$marker %in% cellTypeMarkers), 
                        fdr = case_when(stateChanges2$fdr < 0.05 ~ "significant"), type = "None_VOR"),
              data.frame(TP =cumsum(!stateChanges2Corrected$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges2Corrected$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges2Corrected$fdr < 0.05 ~ "significant"), type = "Corrected_VOR"),
              data.frame(TP =cumsum(!stateChanges3$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges3$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges3$fdr < 0.05 ~ "significant"), type = "None_10x"),
              data.frame(TP =cumsum(!stateChanges3Corrected$marker %in% cellTypeMarkers), 
                        FP = cumsum(stateChanges3Corrected$marker %in% cellTypeMarkers),
                        fdr = case_when(stateChanges3Corrected$fdr < 0.05 ~ "significant"), type = "Corrected_10x"))
  # print(head(df[df$type == "Corrected",]))
  # print(head(df[df$type == "None_BIDC",]))
  # print(tail(df[df$type == "None_BIDC",]))
  # print(head(df[df$type == "None_BIDC",]))

  # slope <- df[nrow(df),]["TP"] / df[nrow(df),]["FP"]
  slope <- nrow(stateChanges %>% filter(!marker %in% cellTypeMarkers))/nrow(stateChanges %>% filter(marker %in% cellTypeMarkers))
  slope <- slope %>% as.numeric
  p1 <- ggplot(df, aes(x = FP, y = TP, colour = type)) + 
        geom_line()+ labs(y = "Cell state marker", x = "Cell type marker") + 
        scale_colour_manual(values = values) +
        geom_abline(intercept = 0, slope = slope, linetype = "dashed", color = "#000000") +
        ggtitle(title)


  p2 <- ggplot(df, aes(x = FP, y = TP, colour = type)) + 
        geom_line() +
        xlim(0,xlim) +
        ylim(0,ylim) +
        labs(y = "Cell state marker", x = "Cell type marker") + 
        scale_colour_manual(values = values) +
        geom_abline(intercept = 0, slope = slope, linetype = "dashed", color = "#000000") +
        ggtitle(title)
  # grid.arrange(p1, p2, nrow=2)
  
  list(p1, p2)
}
```

## Generating cell type specific marker list
```{r}
tmp2 <- colData(XeBreast) %>% as.data.frame %>% select(c("cell_id", "scClassify"))
tmp <- t(assay(XeBreast, "logcounts")) %>% bind_cols(tmp2)
allCellMarkers <- rownames(XeBreast)

calculate_percentages <- function(df) {
  df %>%
    summarise_all(~ mean(. > 0, na.rm = TRUE) * 100)
}

df_list <- split(tmp, f = tmp$scClassify)
percentage_results <- lapply(df_list, calculate_percentages)
result_df <- do.call(rbind, percentage_results)

normality <- list()
for(marker in allCellMarkers) {  
  shapiro <- shapiro.test(result_df[marker] %>% unlist())
  normality[marker] <- shapiro$p.value
}

notNormal <- normality[normality[,1] < 0.05,, drop = FALSE]
```

## Figure 3G
```{r}
testCorrection(cellTypeMarkers = rownames(notNormal),
               stateChanges = stateChangesBID,
               stateChangesCorrected = stateChangesBIDConta,
               stateChanges2 = stateChangesVOR,
               stateChanges2Corrected = stateChangesVORConta,
               stateChanges3 = stateChanges10x,
               stateChanges3Corrected = stateChanges10xConta,
               xlim = 1000,
               ylim = 400,
               title = "BIDCell",
               fdr = FALSE
              )


testCorrection(cellTypeMarkers = c("cd3", "cd4", "cd8", "cd56", "cd11c", "cd68", "cd45", "cd20", "cd11b"),
               stateChanges = superListDistance[[1]],
               stateChangesCorrected = superListDistance[[2]],
               stateChanges2 = superListDistance[[3]],
               stateChanges2Corrected = superListDistance[[4]],
               stateChanges3 = superListDistance[[5]],
               stateChanges3Corrected = superListDistance[[7]],
               xlim = 250,
               ylim = 1000,
               title = "BIDCell",
               fdr = FALSE
              )
```




```{r}
plotStateChanges(schurchSCE,
                 image = "reg002_B", 
                 from = "granulocytes",
                 to = "vasculature",
                 marker = "cd4_pd_1")
```



















